---
title: "MW-EEG project: analysis of behavioral response data"
author: "J.Simola"
date: "2022-09-28"
output: html_notebook
---

```{r}
library(lme4)
library(lmerTest)
library(readr)
library(dplyr)
library(sjPlot) # table functions
library(lattice) # needed for qqmath
library(rcompanion)
library(ggplot2)
```

```{r}
rm(list = ls()) # clear workspace
```

## Load data - open file behavioral.RData 

# Prepare data for analysis 
```{r}
D1 <- D0
D1 <- D1[complete.cases(D1), ] # exclude NAs
D1$run <- factor(D1$run)

D1 <- filter(D1, event != 6) # exclude outlier level 
D1$cond <- factor(D1$event, labels = c("0 back","1 back")) # rename events
levels(D1$cond)

D1$RT <- D1$RT/10 # RTs to ms
D1 <- filter(D1, resp_code != -2) # exclude missing responses

table(D1$resp_code)

# remove RT outliers by 1.5*IQR range
Q <- quantile(D1$RT, probs=c(.25, .75), na.rm = FALSE)
iqr <- IQR(D1$RT)
D1 <- subset(D1, D1$RT > (Q[1] - 1.5*iqr) & D1$RT < (Q[2]+1.5*iqr))

# look at RT distributions
plotNormalHistogram(D1$RT)
qqnorm(D1$RT)
qqline(D1$RT)

D1 <- filter(D1, NT_ord >= 2 & NT_ord <= 6) # exclude NT_ord outliers
hist(D1$NT_ord)

D1
```

# Descriptive results (Table 2)
```{r}
# run: 0 = 1st session (without TPs); 1 = 2nd session (with TPs); 2 = 3rd session (with TPs)
D1 %>% 
  group_by(cond,run) %>% 
  summarize(mean_acc = mean(resp_code),
            sd_acc = sd(resp_code),
            mean_RT = mean(RT),
            sd_RT = sd(RT))

```

## Behavioral analyses ##

# Predicting accuracy using a binomial GLMM - best model found with forward selection heuristic 
```{r}
# NT_ord: elapsed time from the last target trial
acc.0 <- glmer(resp_code ~ 1 + (1 + cond | subj), data=D1, family="binomial")
acc.1 <- glmer(resp_code ~ cond + (1 + cond | subj), data=D1, family="binomial")
acc.2 <- glmer(resp_code ~ cond + run + (1 + cond | subj), data=D1, family="binomial")
acc.3 <- glmer(resp_code ~ cond + run + NT_ord + (1 + cond | subj), data=D1, family="binomial")
acc.4 <- glmer(resp_code ~ cond + NT_ord + (1 + cond | subj), data=D1, family="binomial")
acc.5 <- glmer(resp_code ~ cond * run + (1 + cond | subj), data=D1, family="binomial")
acc.6 <- glmer(resp_code ~ cond * NT_ord + (1 + cond | subj), data=D1, family="binomial")
acc.7 <- glmer(resp_code ~ cond * run * NT_ord + (1 + cond | subj), data=D1, family="binomial") # did not converge

# compare models with likelihood ratio test
anova(acc.0, acc.1, acc.2, acc.3, test = "LRT")
anova(acc.1, acc.4, test = "LRT") 
anova(acc.2, acc.5, test = "LRT") # compare interaction model to that with main effects
anova(acc.4, acc.6, test = "LRT") # compare interaction model to that with main effects
anova(acc.3, acc.7, test = "LRT") # compare interaction model to that with main effects

# acc.2 is the best model
summary(acc.2 <- glmer(resp_code ~ cond + run + (1 + cond | subj), data=D1, family="binomial"))
confint(acc.2, method = "Wald") # create confidence intervals

#investigate session main effect by releveling
D1$run <- relevel(D1$run, 3)
levels(D1$run) # check baseline - remember to level back to the orig ("0","1","2")

summary(acc.2.1 <- glmer(resp_code ~ cond + run + (1 + cond | subj), data=D1, family="binomial"))
confint(acc.2.1, method = "Wald") # create confidence intervals
```


# Predicting RT using an LMM - best model found with forward selection heuristic 
```{r}
rt.0 <- lmer(RT ~ 1 + (1 + cond | subj), data=D1, REML = FALSE)
rt.1 <- lmer(RT ~ cond + (1 + cond | subj), data=D1, REML = FALSE)
rt.2 <- lmer(RT ~ cond + run + (1 + cond | subj), data=D1, REML = FALSE)
rt.3 <- lmer(RT ~ cond + run + NT_ord + (1 + cond | subj), data=D1, REML = FALSE)
rt.4 <- lmer(RT ~ cond + NT_ord + (1 + cond | subj), data=D1, REML = FALSE)
rt.5 <- lmer(RT ~ cond * NT_ord + (1 + cond | subj), data=D1, REML = FALSE)
rt.6 <- lmer(RT ~ cond * run + (1 + cond | subj), data=D1, REML = FALSE)
rt.7 <- lmer(RT ~ cond * run * NT_ord + (1 + cond | subj), data=D1, REML = FALSE)

# compare models with likelihood ratio test
anova(rt.0, rt.1, rt.2, rt.3, test = "LRT")
anova(rt.2, rt.6, test = "LRT") # compare interaction model to that with main effects
anova(rt.4, rt.5, test = "LRT") 
anova(rt.3, rt.7, test = "LRT") 

# rt.5 is the best model
summary(rt.5 <- lmer(RT ~ cond * NT_ord + (1 + cond | subj), data=D1))
confint(rt.5, method = "Wald") # create confidence intervals

residuals <- resid(rt.5) # check that the residuals are normally distributed
hist(residuals)

###### investigate cond x NT_ord interaction #####

# do lmer for a different condition baseline 
D1$cond <- relevel(D1$cond, 2)
levels(D1$cond)

summary(rt.5.1 <- lmer(RT ~ cond * NT_ord + (1 + cond | subj), data=D1))
confint(rt.5.1, method = "Wald") # create confidence intervals
```


# Test the effect of P3 on task performance (accuracy & RTs) 
```{r}
summary(acc.0 <- glmer(resp_code ~ Pz +  (1 + cond | subj), data=D1, family="binomial"))
confint(acc.0, method = "Wald") 

summary(acc.1 <- glmer(resp_code ~ Pz + cond + run + (1 + cond | subj), data=D1, family="binomial"))
confint(acc.1, method = "Wald") 

summary(acc.2 <- glmer(resp_code ~ cond + run + (1 + cond | subj), data=D1, family="binomial"))
confint(acc.2, method = "Wald") 

anova(acc.1, acc.2)

# -------- RT (see Figure 5) --------

summary(rt.0 <- lmer(RT ~ Pz + (1+cond | subj), data=D1))
confint(rt.0, method = "Wald") 

summary(rt.1 <- lmer(RT ~ Pz + cond * NT_ord + (1+cond | subj), data=D1))
confint(rt.1, method = "Wald") 
```


# Write behavioral summary data a file 
```{r}
D1_TP <- D1
D1_TP$run <- as.numeric(levels(D1_TP$run))[D1_TP$run]
D1_TP <- filter(D1_TP, run > 0) # exclude the run without thought probes

beh_summary <- 
  D1_TP %>% 
  group_by(subj, run,cond) %>% 
  summarise(acc = mean(resp_code),
            RT = mean(RT))

beh_summary
write.csv(beh_summary,"/Users/jsimola/Documents/Opetus/Kogn_neuro_harjoituskurssi/R_analysis/data/beh_summary_220906.csv", row.names = FALSE)
```


## Plot (Figure 2)

# Plot accuracy as a function of time and task (Fig 2a)
```{r}
D1.0 <- summarySE(D1, measurevar="resp_code", groupvars=c("NT_ord","cond")) # remember the run the summarySE function below!!
D1.0

setEPS()
postscript("plots/ACC_by_cond_NT_ord.eps")
ggplot(D1.0, aes(x = NT_ord, y = resp_code, color=cond)) +
  geom_errorbar(aes(ymin=resp_code-se, ymax=resp_code+se), width=.1) + 
  scale_color_manual(values=c('darkgrey','black')) +
  ylab("Accuracy") +
  xlab("Elapsed # trials") +
  geom_line() +
  geom_point() +
  theme_void() +
  theme_bw()
```

# Plot RTs as a function of time and task (Fig 2b)
```{r}
D1.0 <- summarySE(D1, measurevar="RT", groupvars=c("NT_ord","cond")) # remember the run the summarySE function below!!
D1.0

setEPS()
postscript("plots/RT_by_cond_NT_ord.eps")
ggplot(D1.0, aes(x = NT_ord, y = RT, color=cond)) +
  geom_errorbar(aes(ymin=RT-se, ymax=RT+se), width=.1) + 
  scale_color_manual(values=c('darkgrey','black')) +
  ylab("RT (ms)") +
  xlab("Elapsed # trials") +
  geom_line() +
  geom_point() +
  theme_void() +
  theme_bw()
```


# Custom function to extract standard error (SE) - run this before plotting
```{r}
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}
```
