---
title: "MW-EEG project: Supplementary analysis"
author: "J.Simola"
date: "2022-10-11"
output: html_document
---

# Test for different component rotations 

```{r}
library(lme4)
library(lmerTest)
library(readr)
library(sjPlot) # table functions
library(dplyr)
library(ggplot2)
library(rcompanion) # plot normal histogram
library(emmeans) # follow-up analysis - pairwise comparisons for factors
```

```{r}
rm(list = ls()) # clear workspace
```

## Load data - open file thoughtprobe.RData 

# Prepare data for analysis 
```{r}
TP <- TP[complete.cases(TP), ] # exclude NAs
TP$cond <- factor(TP$event, labels = c("0 back","1 back")) # rename events
TP$run <- factor(TP$run)
TP <- filter(TP, NT_ord >= 2 & NT_ord <= 6)

# remove Pz outliers by 1.5*IQR range
Q <- quantile(TP$Pz, probs=c(.25, .75), na.rm = FALSE)
iqr <- IQR(TP$Pz)
TP <- subset(TP, TP$Pz > (Q[1] - 1.5*iqr) & TP$Pz < (Q[2]+1.5*iqr))

TP
```

# Task effects for oblimin rotated PCA-components (see S1. Principal component analysis (PCA) options)
```{r}
summary(pca1.ob <- lmer(pca_c1_ob ~ cond + (1+cond | subj), data=TP))
confint(pca1.ob, method = "Wald")

summary(pca2.ob <- lmer(pca_c2_ob ~ cond + (1+cond | subj), data=TP))
confint(pca2.ob, method = "Wald")

summary(pca3.ob <- lmer(pca_c3_ob ~ cond + (1+cond | subj), data=TP))
confint(pca3.ob, method = "Wald")

summary(pca4.ob <- lmer(pca_c4_ob ~ cond + (1 | subj), data=TP))
confint(pca4.ob, method = "Wald")
```

# Predicting the oblimin rotated PCA-component scores using an LMM (see S1. Principal component analysis (PCA) options)
```{r}
TP$PCA <- D1$pca_c1_ob # Note! Select the component here (!!)

m0 <- lmer(Pz ~ PCA + cond + (1 + cond | subj), data=TP, REML = FALSE)
m1 <- lmer(Pz ~ PCA + cond + NT_ord + (1 + cond | subj), data=TP, REML = FALSE)
m2 <- lmer(Pz ~ PCA + cond + run + (1 + cond | subj), data=TP, REML = FALSE)
m3 <- lmer(Pz ~ PCA * cond * NT_ord + (1 + cond | subj), data=TP, REML = FALSE)

# compare models with likelihood ratio test
anova(m0, m1, m2, test = "LRT")
anova(m1, m3, test = "LRT") # compare interaction models to those with main effects

# pca_c1_ob: m2 is the best model 
summary(m2 <- lmer(Pz ~ PCA + cond + NT_ord + (1 + cond | subj), data=TP))
confint(m2, method = "Wald") # create confidence intervals   
```

# Task effects for the three-component PCA solution (see S2. Three-component PCA solution)
```{r}
summary(pca1.ob <- lmer(pca_c1_v3 ~ cond + (1+cond | subj), data=TP))
confint(pca1.ob, method = "Wald")

summary(pca2.ob <- lmer(pca_c2_v3 ~ cond + (1+cond | subj), data=TP))
confint(pca2.ob, method = "Wald")

summary(pca3.ob <- lmer(pca_c3_v3 ~ cond + (1+cond | subj), data=TP))
confint(pca3.ob, method = "Wald")
```

# Predicting the component scores for the three-component PCA solution (see S2. Three-component PCA solution)
```{r}
TP$PCA <- D1$pca_c1_v3 # Note! Select the component here (!!)

m0 <- lmer(Pz ~ cond + (1 + cond | subj), data=TP, REML = FALSE)
m1 <- lmer(Pz ~ PCA + cond + (1 + cond | subj), data=TP, REML = FALSE)
m2 <- lmer(Pz ~ PCA + cond + run + (1 + cond | subj), data=TP, REML = FALSE)
m3 <- lmer(Pz ~ PCA + cond + NT_ord + (1 + cond | subj), data=TP, REML = FALSE)
m4 <- lmer(Pz ~ PCA * cond * NT_ord + (1 + cond | subj), data=TP, REML = FALSE)

# compare models with likelihood ratio test
anova(m0, m1, m2, test = "LRT")
anova(m1, m3, test = "LRT")
anova(m3, m4, test = "LRT") # compare interaction models to those with main effects

# pca_c1_v3: m3 is the best model 
summary(m3 <- lmer(Pz ~ PCA + cond + NT_ord + (1 + cond | subj), data=TP))
confint(m3, method = "Wald") # create confidence intervals   
```

